---
title: "Lab 05 - La Quinta is Spanish for next to Denny's, Pt. 2"
author: "Noah Booker"
date: "3/26/25"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
#library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1
Filter the Denny’s data frame for Alaska (AK) and save the result as dn_ak. How many Denny’s locations are there in Alaska?
```{r ex1_a}
load("~/Documents/Documents/Studies/School/WFU/DS4P/Lab_5/data/dennys.rda")
load("~/Documents/Documents/Studies/School/WFU/DS4P/Lab_5/data/laquinta.rda")
dn <- dennys
lq <- laquinta
dn_ak <- dn %>%
  filter(state == "AK")
nrow(dn_ak)
```

There are three Denny's locations in Alaska.

Now, do the same for La Quinta data frame for Alaska (AK) and save the result as lq_ak. How many La Quinta locations are there in Alaska?

```{r ex1_b}
lq_ak <- lq %>% 
  filter(state == "AK")
nrow(lq_ak)
```

There are two La Quinta locations in Alaska.


### Exercise 2
Next we will be calculating the distance between all Denny’s and all La Quinta locations in Alaska.

How many pairings are there between all Denny’s and all La Quinta locations in Alaska, i.e., how many distances do we need to calculate between the locations of these establishments in Alaska?

3 x 2 = 6


Calculate the number of pairings using the data frames you have already made.

```{r combos}
#Why would I need code for this? dn has 2 observations, lq has 3. Number of pairings is 2 x 3 = 6.
```


### Exercise 3
In order to calculate these distances, we need to first restructure our data to pair the Denny’s and La Quinta locations. To do so, we will join the two data frames.
```{r full_join}
dn_lq_ak <- full_join(dn_ak, lq_ak, 
                      by = "state")
dn_lq_ak
```


### Exercise 4
How many observations are in the joined dn_lq_ak data frame? What are the names of the variables in this data frame?

We see in the tibble above that there are 6 observations on 11 variables. One variable, the one that was common between both datasets, is state. The others are a .x and .y version of address, city, zip, longitude, and latitude——the .x versions coming from the dn dataset, indicating Denny's locations, and the .y versions coming from the lq dataset, indicating La Quinta locations.


### Exercise 5
What function from the tidyverse do we use to add a new variable to a data frame while keeping the existing variables?

mutate( ).

```{r Haversine}
haversine <- function(long1, lat1, long2, lat2, round = 3) {
  # convert to radians
  long1 <- long1 * pi / 180
  lat1 <- lat1 * pi / 180
  long2 <- long2 * pi / 180
  lat2 <- lat2 * pi / 180

  R <- 6371 # Earth mean radius in km

  a <- sin((lat2 - lat1) / 2)^2 + cos(lat1) * cos(lat2) * sin((long2 - long1) / 2)^2
  d <- R * 2 * asin(sqrt(a))

  return(round(d, round)) # distance in km
}
```


### Exercise 6
Calculate the distances between all pairs of Denny’s and La Quinta locations and save this variable as distance. Make sure to save this variable in THE dn_lq_ak data frame, so that you can use it later.

```{r distances}
dn_lq_ak <- dn_lq_ak %>% 
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 3))
```

### Exercise 7
Calculate the minimum distance between a Denny’s and La Quinta for each Denny’s location. To do so we group by Denny’s locations and calculate a new variable that stores the information for the minimum distance.

```{r min_distance}
dn_lq_ak_mindist <- dn_lq_ak %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

dn_lq_ak_mindist
```

### Exercise 8
Describe the distribution of the distances of Denny’s and the nearest La Quinta locations in Alaska. Also include an appropriate visualization and relevant summary statistics.

Well, describing the distribution is pretty easy since there are only three data points. There are three Denny's locations with distances to the nearest La Quinta of 5.20, 2.04, and 6 km.

```{r}
#Here is what I came up with on my own (except that Claude.ai helped me figure out I needed to put x = "" to get geom_point( ) to work).
dn_lq_ak_mindist %>% 
  ggplot(aes(x = "", y = closest)) +
  geom_boxplot() +
  geom_point()

#Claude helped much more with this version.
# Create a box plot with labeled data points
ggplot(dn_lq_ak_mindist, aes(x = "", y = closest)) +
  geom_boxplot(width = 0.5, fill = "lightblue", alpha = 0.7) +
  geom_point(size = 3, color = "blue", alpha = 0.7) +
  # Add text labels for each point
  geom_text(aes(x = 1.15, label = address.x), hjust = 0, size = 3) +
  labs(title = "Distances from Denny's to Nearest La Quinta in Alaska",
       y = "Distance (km)",
       x = "") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

#Relevant summary statistics.
dn_lq_ak_mindist %>%
  pull(closest) %>%
  summary()
```

### Exercise 9
Repeat the same analysis for North Carolina: (i) filter Denny’s and La Quinta Data Frames for NC, (ii) join these data frames to get a complete list of all possible pairings, (iii) calculate the distances between all possible pairings of Denny’s and La Quinta in NC, (iv) find the minimum distance between each Denny’s and La Quinta location, (v) visualize and describe the distribution of these shortest distances using appropriate summary statistics.

```{r NC}
dn_nc <- dn %>%
  filter(state == "NC")

lq_nc <- lq %>% 
  filter(state == "NC")

dn_lq_nc <- full_join(dn_nc, lq_nc, 
                      by = "state")

dn_lq_nc <- dn_lq_nc %>% 
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 3))

dn_lq_nc_mindist <- dn_lq_nc %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

dn_lq_nc_mindist %>% 
  ggplot(aes(x = "", y = closest)) +
    geom_boxplot(width = 0.5, fill = "lightblue", alpha = 0.7) +
    geom_point(size = 3, color = "blue", alpha = 0.7) +
    labs(title = "Distances from Denny's to Nearest La Quinta in North Carolina",
       y = "Distance (km)",
       x = "") +
    theme_minimal() +
    theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

dn_lq_nc_mindist %>%
  pull(closest) %>%
  summary()
```

### Exercise 10
Repeat the same analysis for Texas.
```{r TX}
dn_tx <- dn %>%
  filter(state == "TX")

lq_tx <- lq %>% 
  filter(state == "TX")

dn_lq_tx <- full_join(dn_tx, lq_tx, 
                      by = "state")

dn_lq_tx <- dn_lq_tx %>% 
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 3))

dn_lq_tx_mindist <- dn_lq_tx %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

ggplot(dn_lq_tx_mindist, aes(x = "", y = closest)) +
  geom_boxplot(width = 0.5, fill = "lightblue", alpha = 0.7) +
  geom_point(size = 3, color = "blue", alpha = 0.7) +
  labs(title = "Distances from Denny's to Nearest La Quinta in Texas",
       y = "Distance (km)",
       x = "") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

dn_lq_tx_mindist %>%
  pull(closest) %>%
  summary()
```


### Exercise 11
Repeat the same analysis for a state of your choosing, different than the ones we covered so far.
```{r LA}
dn_la <- dn %>%
  filter(state == "LA")

lq_la <- lq %>% 
  filter(state == "LA")

dn_lq_la <- full_join(dn_la, lq_la, 
                      by = "state")

dn_lq_la <- dn_lq_la %>% 
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 3))

dn_lq_la_mindist <- dn_lq_la %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

ggplot(dn_lq_la_mindist, aes(x = "", y = closest)) +
  geom_boxplot(width = 0.5, fill = "lightblue", alpha = 0.7) +
  geom_point(size = 3, color = "blue", alpha = 0.7) +
  labs(title = "Distances from Denny's to Nearest La Quinta in Louisiana",
       y = "Distance (km)",
       x = "") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

dn_lq_la_mindist %>%
  pull(closest) %>%
  summary()
```

### Exercise 12
Among the states you examined, where is Mitch Hedberg’s joke most likely to hold true? Explain your reasoning.

```{r summaries}
dn_lq_ak_mindist %>%
  pull(closest) %>%
  summary()
dn_lq_nc_mindist %>%
  pull(closest) %>%
  summary()
dn_lq_tx_mindist %>%
  pull(closest) %>%
  summary()
dn_lq_la_mindist %>%
  pull(closest) %>%
  summary()
```

Based on the summary statistics, it seems to me that the states that are in competition are Alaska, Texas, and Louisiana (North Carolina seems way off from what we would expect by looking at the other states). Alaska has the lowest mean, followed by Louisian, and then Texas. However, to make better comparisons, I'll compare the boxplots.

```{r plotting_to_compare}
dn_lq_ak_mindist <- dn_lq_ak_mindist %>% 
  mutate(state = "Alaska")

dn_lq_tx_mindist <- dn_lq_tx_mindist %>% 
  mutate(state = "Texas")

dn_lq_la_mindist <- dn_lq_la_mindist %>% 
  mutate(state = "Louisiana")

dn_lq_mindist_ak_tx <- full_join(
  dn_lq_ak_mindist,
  dn_lq_tx_mindist)

dn_lq_mindist_ak_tx_la <- full_join(
  dn_lq_mindist_ak_tx,
  dn_lq_la_mindist)
  
ggplot(dn_lq_mindist_ak_tx_la, aes(x = "", y = closest)) +
  geom_boxplot(width = 0.5, fill = "lightblue", alpha = 0.7) +
  geom_point(size = 2, color = "blue", alpha = 0.7) +
  facet_wrap(~state) +
  labs(title = "Distances from Denny's to Nearest La Quinta",
       y = "Distance (km)",
       x = "") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

#Replace geom_point with geom_jitter to get a better sense of the number of points on the texas plot.
ggplot(dn_lq_mindist_ak_tx_la, aes(x = "", y = closest)) +
  geom_boxplot(width = 0.5, fill = "lightblue", alpha = 0.7) +
  geom_jitter(width = 0.03, height = 0, size = 2, color = "blue", alpha = 0.7) +
  facet_wrap(~state) +
  labs(title = "Distances from Denny's to Nearest La Quinta",
       y = "Distance (km)",
       x = "") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```

Based on a comparison of the boxplots, it seems that the joke holds most true in Texas. It has the lowest median and by far the most numerous Denny's locations that have a distance to the closest La Quinta that is below ~3 km.