---
title: "Lab 05 - La Quinta is Spanish for next to Denny's, Pt. 2"
author: "Noah Booker"
date: "3/26/25"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
#library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1
Filter the Denny’s data frame for Alaska (AK) and save the result as dn_ak. How many Denny’s locations are there in Alaska?
```{r ex1_a}
load("~/Documents/Documents/Studies/School/WFU/DS4P/Lab_5/data/dennys.rda")
load("~/Documents/Documents/Studies/School/WFU/DS4P/Lab_5/data/laquinta.rda")
dn <- dennys
lq <- laquinta
dn_ak <- dn %>%
  filter(state == "AK")
nrow(dn_ak)
```

There are three Denny's locations in Alaska.

Now, do the same for La Quinta data frame for Alaska (AK) and save the result as lq_ak. How many La Quinta locations are there in Alaska?

```{r ex1_b}
lq_ak <- lq %>% 
  filter(state == "AK")
nrow(lq_ak)
```

There are two La Quinta locations in Alaska.


### Exercise 2

Remove this text, and add your answer for Exercise 1 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 3

...

### Exercise 4

...

### Exercise 5

...

### Exercise 6

...

Add exercise headings as needed.
